Association between trauma severity models and opportunities for improvement:

### 2022-12-22 manuscript.Rmd
Minor edits

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->
*Background:* Morbidity and mortality review, which aims to discover opportunities for improvement (OFI) is a cornerstone of trauma quality improvement worldwide. The process of case selection and review is time-consuming, leading many hospitals to use audit filters to automate case selection. Trauma severity models, which estimate probability of survival, have been proposed as one such audit filter. The ability of these models to predict opportunities for improvement (OFI) is not well understood. Two studies have found that they perform poorly at predicting preventable deaths, but as of yet there is no published literature that analyses the performance of these models across the board and not only in deaths. *Aim:* To evaluate how the commonly used models TRISS, NORMIT and PS are associated with, and to what extent they can predict, OFI. *Methods:* This is a retrospective registry-based cohort study at Karolinska University Hospital covering the period 2014-2021. Logistic regression was used to estimate the association between the models and OFI. Receiver operating characteristic curves with the corresponding area under the curve (AUC) were used to assess discrimination and the Integrated Calibration Index was used to assess calibration. *Results:* 5320 participants were included from a seven-year period. Results were similar for all models.  Lower probability of survival was associated with higher probability of OFI, with TRISS having the strongest association (derived OR: `r t['or']`). Predictive performance and calibration was poor across the board. TRISS had the best AUC of `r t['auc']`. NORMIT had the best calibration with an ICI of `r n['ici']`. *Conclusions:* Although low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI. The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.

@@

Efforts to advance the state of trauma care have been ongoing for many years. Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi]. QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference. These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case.

One area that is understudied is the use of trauma severity models as audit filters in case selection for M&M conferences. These models are algorithms that estimate the probability of survival. It has been proposed that using such models as audit filters to automatically flag cases for review could improve the effectiveness of M&M conferences while also reducing their resource burden [@brohi16; @prevdeathsghorbani ]. Many hospitals have already implemented this approach [@brohi16]; however, the ability of trauma severity models to predict OFI is not well understood and will be investigated in this paper.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

>>>>

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->
*Background:* Morbidity and mortality review, which aims to discover opportunities for improvement (OFI) is a cornerstone of trauma quality improvement worldwide. The process of case selection and review is time-consuming, leading many hospitals to use audit filters to automate case selection. Trauma severity models, which estimate probability of survival, have been proposed as one such audit filter. Two studies have found that these models perform poorly at predicting preventable deaths, but the ability of these models to predict opportunities for improvement (OFI) is not well understood. *Aim:* To evaluate how the commonly used models TRISS, NORMIT and PS are associated with, and to what extent they can predict, OFI. *Methods:* This is a retrospective registry-based cohort study at Karolinska University Hospital covering the period 2014-2021. Logistic regression was used to estimate the association between the model scores and OFI. Predictive performance was measured using receiver operating characteristic curves with the corresponding area under the curve (AUC) to assess discrimination and the Integrated Calibration Index to assess calibration. *Results:* 5320 participants were included from a seven-year period. Results were similar for all models.  Lower probability of survival was associated with higher probability of OFI, with TRISS having the strongest association (derived OR: `r t['or']`). All model scores had poor predictive performance. TRISS had the best AUC of `r t['auc']`. NORMIT had the best calibration with an ICI of `r n['ici']`. *Conclusions:* Although low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI. The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.

@@

Efforts to advance the state of trauma care have been ongoing for many years. Today many trauma centres have implemented quality improvement (QI)<!-- Do you really need to abbreviate quality improvement? It's quite short as it is and you don't use it too much--> programs, and it has been shown in the US that these programs improve outcomes [@ustqi]. QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference. These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case.

One area that is understudied is the use of trauma severity models as audit filters in case selection for M&M conferences. These models are algorithms that estimate the probability of survival. It has been proposed that using such models as audit filters to automatically flag cases for review could improve the effectiveness of M&M conferences while also reducing their resource burden [@brohi16; @prevdeathsghorbani]. Many hospitals have already implemented this approach [@brohi16]; however, the ability of trauma severity models to predict OFI is not well understood and will be investigated in this paper.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority in treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].
